```{r setupdata, include=FALSE}
rm(list = ls()) ; invisible(gc()) ; set.seed(42)
library(knitr)
library(kableExtra)
if(knitr:::is_html_output()) options(knitr.table.format = "html") 
if(knitr:::is_latex_output()) options(knitr.table.format = "latex") 
library(tidyverse)
library(ggfortify)
theme_set(bayesplot::theme_default())
opts_chunk$set(
  echo = F, message = F, warning = F, fig.height = 6, fig.width = 8,
  cache = T, cache.lazy = F)
```

# Data

This chapter introduce and explore explanatory variables correlations and relations with response variables.


## Variables 

* **date**: date DD/MM/YYYYY
* **time**: time HH
* **plot**: plot
* **id**: tree 
* **height**: tree height in m
* **d1**: tree diameter at the base in mm
* **d2**: second tree diameter at the the base orthogonal to d1 in mm
* **diameter**: mean tree diameter in mm ($\frac{d1+d2}{2}$)
* **branches**: number of branches in the tree
* **fruits**: number of fruits in the tree
* **flowers**:  number of flowers in the tree
* **pflower**: presence of flower in the tree (0/1) 
* **nci1**: distance to the closest tree in cm
* **nci2**: distance to the second closest tree in cm
* **nci3**: distance to the third closest tree in cm
* **nci**: mean distance to the three closest trees in cm
* **density**: plot tree density in $trees.m^2$
* **age**: maximum age in months
* **shore**: distance from shore in m. *Distance from the shore is potentially correlated with two important non-measured variables: 1) age since the precise location in the mud-bank became available to the seed rain (nearer plots are likely older than further plots, though not necessarily); and 2) the amount of hours the sea is above that point (nearer plots are likely to be submerged fewer hours).* 
* **soil**: qualitative approximate from mostly sandy (1) to deeply muddy (4)
* **openness**: qualitative approximate of light that reached the soil from a not densely covered/ lot of light (1) to very densely covered and very little light reaching the bottom (4)

```{r}
data <- read_tsv("data/mangrove_data.txt") %>% 
  dplyr::rename(plot = `Placettes`,
               id = `ID`,
               height = `Hauteur(m)`,
               d1 = `Diametre1(mm)`,
               d2 = `Diametre2(mm)`,
               diameter = `Moyennediam(mm)`,
               branches = `Branches(N)`,
               fruits = `Fruits(N)`,
               flowers = `Fleurs(N)`,
               pflower= `Presencefleurs(1/0)`,
               nci1 = `Competition_1(cm)`,
               nci2 = `Competition_2(cm)`,
               nci3 = `Competition_3(cm)`,
               nci = `Moyennecompetition(cm)`,
               density = `Densite(trees/m2)`,
               date = `Date`,
               time = `Heure`,
               soil = `Typesol`,
               openness = `Ouverture`,
               age = `Agemax(months)`,
               shore = `Distance(fromShore)`) %>% 
  mutate(time = as.numeric(gsub("h", "", time)))
```
## Model types

* **Lognormal regressions**: `height` and `diameter` are positive continuous measurements and followed a lognormal distribution
* **Poisson exponential regressions**:  `branches`, `fruits` and `flowers` are count data and followed a Poisson distribution
* **Bernoulli logistic regressions**: `pflower` is binary and followed a Bernoulli distribution

```{r resphist}
dplyr::select(data, 
              height, diameter, branches, fruits, flowers, pflower) %>% 
  reshape2::melt() %>% 
  ggplot(aes(value)) +
  geom_histogram(aes(y = ..density..)) +
  geom_density() +
  facet_wrap(~ variable, scales = "free")
```



## Co-variations

Figures \@ref(fig:cor) and \@ref(fig:pca) showed variables co-variations among them.
There are a lot of positive and negative co-variations, as indicated by Niklas Tysklind preliminary results.
Growth and (height, diameter, and branches) are positively correlated to reproduction (fruits, flower, pflowers),
which may simplify the analyses if we find similar results for the two performance traits.
NCI was correlated to both growth and reproduction.
Soil was negatively correlated to growth.
Openness was positively correlated to growth.
Due to the high correlation between density and age ($r=0.74$, \@ref(fig:cor)),
we will drop age from further analyses (density is more accurate).

```{r cor, fig.cap="Variables correlations. Pearson's r below the diagonal and corresponding color above the diagonal."}
dplyr::select(data, 
              height, diameter, branches, fruits, flowers, pflower, nci, density, age, shore, soil, openness) %>% 
  na.omit() %>%
  cor() %>% 
  corrplot::corrplot.mixed()
```

```{r pca, fig.cap="Principal component analysis of variables."}
autoplot(princomp(~ height + diameter + branches + fruits + flowers + pflower + nci + density + soil + openness + age + shore,
                  data = data, cor = T),
         col = "lightgrey", alpha = 0.5,
         loadings.colour = "black",
         loadings.label.size = 6, loadings.label.repel	= T,
         loadings.label.colour = 'black', loadings.label.vjust = 1.1,
         loadings = T, loadings.label = T) +
  geom_hline(aes(yintercept = 0), col = 'black', linetype = "dotted") +
  geom_vline(aes(xintercept = 0), col = 'black', linetype = "dotted")
```

## Relations

We explored relations among variables using:

* **Response variables**: 
    * **Growth**: height, diameter, and branches 
    * **Reproduction**: fruits, flowers and pflower
* **Explanatory variables**: 
    * **Continuous**: nci, density, age, and shore
    * **Qualitative**: soil, and openness 
* **Random effects**: plot


Figures \@ref(fig:lmplotcontinuous), \@ref(fig:lmplotcategorial) and \@ref(fig:oob) and table \@ref(tab:lmtab) showed explanatory variables relation to response variables.
All explanatory variables had a significant linear effect on all response variables (Fig. \@ref(fig:oob)).
Explanatory variables explained 8 to 15% of variation in reproductive traits and 41 to 63% of variation in growth traits (Tab. \@ref(tab:lmtab)).

```{r lmplotcontinuous, fig.cap="Relations for continuous explanatory variables."}
dplyr::select(data,
              id,
              height, diameter, branches, fruits, flowers, pflower, # response variables
              nci, density, shore # explanatory variables continuous
              ) %>% 
  reshape2::melt(id.vars = c("id", "height", "diameter", "branches", "fruits", "flowers", "pflower"),
                 variable.name = "explanatory", value.name = "explanatory.value") %>% 
    reshape2::melt(id.vars = c("id", "explanatory", "explanatory.value"),
                 variable.name = "response", value.name = "response.value") %>% 
  ggplot(aes(x = explanatory.value, y = response.value)) +
  geom_point() +
  facet_grid(response ~ explanatory, scales = "free") +
  geom_smooth() +
  xlab("Explanatory") + ylab("Response")
```

```{r lmplotcategorial, fig.cap="Relations for categorial explanatory variables."}
dplyr::select(data,
              id,
              height, diameter, branches, fruits, flowers, # response variables
              soil, openness # explanatory variables factor and random
              ) %>% 
  reshape2::melt(id.vars = c("id", "height", "diameter", "branches", "fruits", "flowers"),
                 variable.name = "explanatory", value.name = "explanatory.value") %>% 
    reshape2::melt(id.vars = c("id", "explanatory", "explanatory.value"),
                 variable.name = "response", value.name = "response.value") %>% 
  ggplot(aes(x = as.factor(explanatory.value), y = log(response.value))) +
  geom_boxplot() +
  facet_grid(response ~ explanatory, scales = "free") +
  xlab("Explanatory") + ylab("Response")
```
```{r oob, fig.cap="Variable importance calculated with an out-of-bag method applied on a random forest. Variable importance is represented by the mean increase in mean square error (MSE)."}
dplyr::select(data,
              id,
              height, diameter, branches, fruits, flowers, pflower, # response variables
              nci, density, soil, openness, shore # explanatory variables
              ) %>% 
    reshape2::melt(id.vars = c("id", "nci", "density", "soil", "openness", "shore"),
                   variable.name = "response") %>% 
  group_by(response) %>%
  mutate(soil = as.factor(soil), openness = as.factor(openness)) %>% 
  do(oob= randomForest::randomForest(value ~ nci + density + soil + openness + shore,
                                     data = .,  na.action = na.omit,
                                     ntree = 1000, replace = F , importance = T,
                                     do.trace = F, keep.forest = T, keep.inbag = T)$importance %>% 
         as.data.frame() %>%
  rownames_to_column("explanatory") %>%
  dplyr::rename(MSEinc = "%IncMSE")) %>% 
  unnest(oob) %>% 
  ggplot(aes(explanatory, MSEinc)) +
  geom_col() +
  coord_flip() +
  xlab("") + ylab("Mean increase in mean square error.") +
  facet_wrap(~ response, scales = "free")
```


## Conclusion

We already found a lot of variation explained, but there are still a lot of unexplained variation, especially for reproduction. 
In the final model, we can add plot, date and hours random effects to remove plot, climate and weather effects from residual variation,
and get closer to the individual variation.
Then we may be able to identify outliers with high/slow growth/reproduction.

