```{r setupdata, include=FALSE}
rm(list = ls()) ; invisible(gc()) ; set.seed(42)
library(knitr)
library(kableExtra)
if(knitr:::is_html_output()) options(knitr.table.format = "html") 
if(knitr:::is_latex_output()) options(knitr.table.format = "latex") 
library(tidyverse)
library(ggfortify)
theme_set(bayesplot::theme_default())
opts_chunk$set(
  echo = F, message = F, warning = F, fig.height = 6, fig.width = 8,
  cache = T, cache.lazy = F)
```

# Data

This chapter introduce and explore explanatory variables correlations and relations with response variables.


## Variables 

*Renamed.*

* **plot**: plot
* **id**: tree 
* **height**: tree height in m
* **d1**: ? in mm
* **d2**: ? in mm
* **diameter**: tree diameter in mm
* **branches**: number of branches in the tree
* **fruits**: number of fruits in the tree
* **flowers**:  number of flowers in the tree
* **pflower**: presence of flower in the tree (0/1) 
* **nci1**: distance to the closest tree in cm
* **nci2**: distance to the second closest tree in cm
* **nci3**: distance to the third closest tree in cm
* **nci**: mean distance to the three closest trees in cm
* **density**: plot tree density in $trees.m^2$
* **date**: date DD/MM/YYYYY
* **time**: time HH
* **soil**: type of soil (1-4 ?)
* **openness**: ?
* **age**: maximum age in months
* **shore**: distance from shore in m  

```{r}
data <- read_tsv("data/mangrove_data.txt") %>% 
  dplyr::rename(plot = `Placettes`,
               id = `ID`,
               height = `Hauteur(m)`,
               d1 = `Diametre1(mm)`,
               d2 = `Diametre2(mm)`,
               diameter = `Moyennediam(mm)`,
               branches = `Branches(N)`,
               fruits = `Fruits(N)`,
               flowers = `Fleurs(N)`,
               pflower= `Presencefleurs(1/0)`,
               nci1 = `Competition_1(cm)`,
               nci2 = `Competition_2(cm)`,
               nci3 = `Competition_3(cm)`,
               nci = `Moyennecompetition(cm)`,
               density = `Densite(trees/m2)`,
               date = `Date`,
               time = `Heure`,
               soil = `Typesol`,
               openness = `Ouverture`,
               age = `Agemax(months)`,
               shore = `Distance(fromShore)`) %>% 
  mutate(time = as.numeric(gsub("h", "", time)))
```

## Co-variations

Figures \@ref(fig:cor) and \@ref(fig:pca) showed variables co-variations among them.
There are a lot of positive and negative co-variations, as indicated by Niklas Tysklind preliminary results.
Growth and (height, diameter, and branches) are positively correlated to reproduction (fruits, flower, pflowers),
which may simplify the analyses if we find similar results for the two performance traits.
NCI was correlated to both growth and reproduction.
Soil was negatively correlated to growth.
Openness was positively correlated to growth.

```{r cor, fig.cap="Variables correlations. Pearson's r below the diagonal and corresponding color above the diagonal."}
dplyr::select(data, 
              height, diameter, branches, fruits, flowers, pflower, nci, density, soil, openness, age, shore) %>% 
  na.omit() %>%
  cor() %>% 
  corrplot::corrplot.mixed()
```

```{r pca, fig.cap="Principal component analysis of variables."}
autoplot(princomp(~ height + diameter + branches + fruits + flowers + pflower + nci + density + soil + openness + age + shore,
                  data = data, cor = T),
           loadings.label.size = 6,
           loadings.label.colour = 'black', loadings.label.vjust = 1.1,
           loadings = T, loadings.label = T, loadings.colour = 'black') +
  geom_hline(aes(yintercept = 0), col = 'black', linetype = "dotted") +
  geom_vline(aes(xintercept = 0), col = 'black', linetype = "dotted")
```

## Relations

We explored relations among variables using:

* **Response variables**: height, diameter, and branches for growth and fruits, flowers and pflower for reproduction
* **Explanatory variables**: nci, density, soil, openness, age, and shore
* **Random effects**: plot, date, and hour


Figures \@ref(fig:lmplot) and \@ref(fig:oob) and table \@ref(tab:lmtab) showed explanatory variables relation to response variables.
All explanatory variables had a significant linear effect on all response variables (**unexpected !**).
Openness, competition, and distance to the shore were the most important variables to describe growth and reproduction (Fig. \@ref(fig:oob)).
Explanatory variables explained 7 to 15% of variation in reproductive traits and 40 to 58% of variation in growth traits (Tab. \@ref(tab:lmtab)).

```{r lmplot, fig.cap="Linear regressions per explanatory variables."}
dplyr::select(data,
              id,
              height, diameter, branches, fruits, flowers, pflower, # response variables
              nci, density, soil, openness, age, shore # explanatory variables
              ) %>% 
  reshape2::melt(id.vars = c("id", "height", "diameter", "branches", "fruits", "flowers", "pflower"),
                 variable.name = "explanatory", value.name = "explanatory.value") %>% 
    reshape2::melt(id.vars = c("id", "explanatory", "explanatory.value"),
                 variable.name = "response", value.name = "response.value") %>% 
  ggplot(aes(x = explanatory.value, y = response.value)) +
  geom_point() +
  facet_grid(response ~ explanatory, scales = "free") +
  geom_smooth(method = "lm") +
  xlab("Explanatory") + ylab("Response")
```

```{r lmtab}
dplyr::select(data,
              id,
              height, diameter, branches, fruits, flowers, # response variables
              nci, density, soil, openness, age, shore # explanatory variables
              ) %>% 
  reshape2::melt(id.vars = c("id", "nci", "density", "soil", "openness", "age", "shore")) %>% 
  group_by(variable) %>%
  mutate(soil = as.factor(soil)) %>% 
  do(lm = broom::glance(lm(value ~ nci + density + soil + openness + age + shore, data = .))) %>% 
  unnest(lm) %>% 
  dplyr::select(variable, r.squared, adj.r.squared) %>% 
  kable(col.names = c("Response", "$R^2$", "$R^2_{adj}$"),
        caption = "Linear regressions per response variable.")
```

```{r oob, fig.cap="Variable importance calculated with an out-of-bag method applied on a random forest. Variable importance is represented by the mean increase in mean square error (MSE)."}
dplyr::select(data,
              id,
              height, diameter, branches, fruits, flowers, pflower, # response variables
              nci, density, soil, openness, age, shore # explanatory variables
              ) %>% 
    reshape2::melt(id.vars = c("id", "nci", "density", "soil", "openness", "age", "shore"),
                   variable.name = "response") %>% 
  group_by(response) %>%
  mutate(soil = as.factor(soil)) %>% 
  do(oob= randomForest::randomForest(value ~ nci + density + soil + openness + age + shore,
                                     data = .,  na.action = na.omit,
                                     ntree = 1000, replace = F , importance = T,
                                     do.trace = F, keep.forest = T, keep.inbag = T)$importance %>% 
         as.data.frame() %>%
  rownames_to_column("explanatory") %>%
  dplyr::rename(MSEinc = "%IncMSE")) %>% 
  unnest(oob) %>% 
  ggplot(aes(explanatory, MSEinc)) +
  geom_col() +
  coord_flip() +
  xlab("") + ylab("Mean increase in mean square error.") +
  facet_wrap(~ response, scales = "free")
```


## Conclusion

We already found a lot of variation explained, but there are still a lot of unexplained variation, especially for reproduction. 
In the final model, we can add plot, date and hours random effects to remove plot, climate and weather effects from residual variation,
and get closer to the individual variation.
Then we may be able to identify outliers with high/slow growth/reproduction.

