```{r setupmodel, include=FALSE, cache=F}
rm(list = ls()) ; invisible(gc()) ; set.seed(42)
library(knitr)
library(kableExtra)
if(knitr:::is_html_output()) options(knitr.table.format = "html") 
if(knitr:::is_latex_output()) options(knitr.table.format = "latex") 
library(tidyverse)
library(ggfortify)
library(rstan)
library(bayesplot)
theme_set(bayesplot::theme_default())
options(mc.cores = 4)
rstan_options(auto_write = TRUE)
opts_chunk$set(
  echo = F, message = F, warning = F, fig.height = 6, fig.width = 8,
  cache = T, cache.lazy = F)
```

```{r mdata}
data <- read_tsv("data/mangrove_data.txt") %>% 
  dplyr::rename(plot = `Placettes`,
                id = `ID`,
                height = `Hauteur(m)`,
                d1 = `Diametre1(mm)`,
                d2 = `Diametre2(mm)`,
                diameter = `Moyennediam(mm)`,
                branches = `Branches(N)`,
                fruits = `Fruits(N)`,
                flowers = `Fleurs(N)`,
                pflower= `Presencefleurs(1/0)`,
                nci1 = `Competition_1(cm)`,
                nci2 = `Competition_2(cm)`,
                nci3 = `Competition_3(cm)`,
                nci = `Moyennecompetition(cm)`,
                density = `Densite(trees/m2)`,
                date = `Date`,
                time = `Heure`,
                soil = `Typesol`,
                openness = `Ouverture`,
                age = `Agemax(months)`,
                shore = `Distance(fromShore)`) %>% 
  mutate(time = as.numeric(gsub("h", "", time))) %>% 
  dplyr::select(
    id,
    height, diameter, branches,
    fruits, flowers, pflower,
    nci, density, shore,
    soil, openness,
    plot
  ) %>% 
  mutate(plotNum = as.numeric(as.factor(plot))) %>% 
  mutate_at(c("nci", "density", "shore"),
            scale) %>% 
  mutate_at(c("nci", "density", "shore"),
            as.vector) %>% 
  reshape2::melt(id.vars = c("id", "nci", "density", "shore", "soil", "openness", "plot", "plotNum"))
```

# Exploratory models

This chapter develop all the models used to explore growth and reproductive variation linked to ontogeny and environment.
We used following variables:

* **Response variables**: 
    * **Growth**: height (lognormal), diameter (lognormal), and branches (Poisson)
    * **Reproduction**: fruits (Poisson), flowers (Poisson), and pflower (Bernoulli)
* **Explanatory variables**: 
    * **Continuous**: nci, density, and shore
    * **Qualitative**: soil, and openness 
* **Random effects**: plot

Most variation was related to openness, soil and neighbours (Fig. \@ref(fig:varpart)). 
Nevertheless, the residual showed a high variation  (Fig. \@ref(fig:varpart)) and structure for `pflower` (Fig. \@ref(fig:residuals)).

```{r varpart, fig.cap="Variance partitionning for all models."}
lm.res <- filter(data, variable %in% c("height", "diameter")) %>% 
  mutate(value = log(value)) %>% 
  group_by(variable) %>% 
  do(varpart = lm(value ~ nci + density + shore + as.factor(soil) + as.factor(openness), .) %>% 
       anova() %>% 
       as.data.frame() %>% 
       rownames_to_column("parameter")) %>% 
  unnest(varpart) %>% 
  group_by(variable) %>% 
  mutate(TotSq = sum(`Sum Sq`)) %>% 
  mutate(pctNum = round(`Sum Sq` / TotSq * 100)) %>% 
  mutate(pctTxt = paste0(pctNum, "%")) %>%
  mutate(parameter = ifelse(parameter == "Residuals", "residuals", parameter)) %>% 
  dplyr::select(variable, parameter, pctNum, pctTxt)
poisson.res <- filter(data, variable %in% c("fruits", "flowers", "branches")) %>% 
  group_by(variable) %>% 
  do(varpart = glm(value ~ nci + density + shore + as.factor(soil) + as.factor(openness), 
      family = poisson, .) %>% 
       anova() %>% 
       as.data.frame() %>% 
       rownames_to_column("parameter")) %>% 
  unnest(varpart) %>% 
  group_by(variable) %>% 
  mutate(TotDev = max(`Resid. Dev`)) %>% 
  mutate(pctNum = round(Deviance / TotDev * 100)) %>% 
  mutate(pctNum = ifelse(is.na(pctNum), 100 - sum(pctNum, na.rm = T), pctNum)) %>% 
  mutate(parameter = ifelse(parameter == "NULL", "residuals", parameter)) %>% 
  mutate(pctTxt = paste0(pctNum, "%")) %>%  
  dplyr::select(variable, parameter, pctNum, pctTxt)
binomial.res <- filter(data, variable %in% c("pflower")) %>% 
  group_by(variable) %>% 
  do(varpart = glm(value ~ nci + density + shore + as.factor(soil) + as.factor(openness), 
      family = binomial, .) %>% 
       anova() %>% 
       as.data.frame() %>% 
       rownames_to_column("parameter")) %>% 
  unnest(varpart) %>% 
  group_by(variable) %>% 
  mutate(TotDev = max(`Resid. Dev`)) %>% 
  mutate(pctNum = round(Deviance / TotDev * 100)) %>% 
  mutate(pctNum = ifelse(is.na(pctNum), 100 - sum(pctNum, na.rm = T), pctNum)) %>% 
  mutate(parameter = ifelse(parameter == "NULL", "residuals", parameter)) %>% 
  mutate(pctTxt = paste0(pctNum, "%")) %>% 
  dplyr::select(variable, parameter, pctNum, pctTxt)
bind_rows(lm.res, poisson.res, binomial.res) %>% 
  ggplot(aes(x = variable, fill = parameter)) +
  geom_col(aes(y = pctNum)) +
  geom_text(aes(y = pctNum, label = pctTxt), col = "white",
            position = position_stack(vjust = .5)) +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank()) +
  viridis::scale_fill_viridis(expression(sigma^2), discrete = T)
```

```{r residuals, fig.cap="Residual for all models."}
lm.res <- filter(data, variable %in% c("height", "diameter")) %>% 
  mutate(value = log(value)) %>% 
  group_by(variable) %>% 
  do(residuals = lm(value ~ nci + density + shore + as.factor(soil) + as.factor(openness), .) %>% 
       residuals() %>% 
       as.tibble()) %>% 
  unnest(residuals)
poisson.res <- filter(data, variable %in% c("fruits", "flowers", "branches")) %>% 
  group_by(variable) %>% 
  do(residuals = glm(value ~ nci + density + shore + as.factor(soil) + as.factor(openness), 
      family = poisson, .) %>% 
      residuals() %>% 
       as.tibble()) %>% 
  unnest(residuals)
binomial.res <- filter(data, variable %in% c("pflower")) %>% 
  group_by(variable) %>% 
  do(residuals = glm(value ~ nci + density + shore + as.factor(soil) + as.factor(openness), 
                     family = binomial, .) %>% 
       residuals() %>% 
       as.tibble()) %>% 
  unnest(residuals)
bind_rows(lm.res, poisson.res, binomial.res) %>% 
  group_by(variable) %>% 
  ggplot(aes(value))  + 
  geom_histogram(aes(y = ..density..), col = "lightgrey", alpha = 0.5) +
  geom_density() +
  facet_wrap(~ variable, scales = "free") +
  xlab("residuals")
```
