```{r setupfull, include=FALSE, cache=F}
rm(list = ls()) ; invisible(gc()) ; set.seed(42)
library(knitr)
library(kableExtra)
if(knitr:::is_html_output()) options(knitr.table.format = "html") 
if(knitr:::is_latex_output()) options(knitr.table.format = "latex") 
library(tidyverse)
library(ggfortify)
library(rstan)
library(loo)
library(bayesplot)
theme_set(bayesplot::theme_default())
options(mc.cores = 4)
rstan_options(auto_write = TRUE)
opts_chunk$set(
  echo = F, message = F, warning = F, fig.height = 6, fig.width = 8,
  cache = T, cache.lazy = F)
```

# Hierarchical full

```{r mdatafull}
data <- read_tsv("data/mangrove_data.txt") %>% 
  dplyr::rename(plot = `Placettes`,
                id = `ID`,
                height = `Hauteur(m)`,
                d1 = `Diametre1(mm)`,
                d2 = `Diametre2(mm)`,
                diameter = `Moyennediam(mm)`,
                branches = `Branches(N)`,
                fruits = `Fruits(N)`,
                flowers = `Fleurs(N)`,
                pflower= `Presencefleurs(1/0)`,
                nci1 = `Competition_1(cm)`,
                nci2 = `Competition_2(cm)`,
                nci3 = `Competition_3(cm)`,
                nci = `Moyennecompetition(cm)`,
                density = `Densite(trees/m2)`,
                date = `Date`,
                time = `Heure`,
                soil = `Typesol`,
                openness = `Ouverture`,
                age = `Agemax(months)`,
                shore = `Distance(fromShore)`) %>% 
  mutate(time = as.numeric(gsub("h", "", time))) %>% 
  dplyr::select(
    id,
    flowers, fruits,
    height, nci, density, plot
  ) %>% 
  mutate(plotNum = as.numeric(as.factor(plot))) %>% 
  mutate_at(c("nci", "density"),
            scale) %>% 
  mutate_at(c("nci", "density"),
            as.vector)
mdata_flowers <- list(
  I = nrow(data),
  P = max(data$plotNum),
  y = data$flowers,
  height = data$height,
  nci = data$nci,
  density = data$density,
  plot = data$plotNum
)
mdata_fruits <- list(
  I = nrow(data),
  P = max(data$plotNum),
  y = data$fruits,
  height = data$height,
  nci = data$nci,
  density = data$density,
  plot = data$plotNum
)
```

## Baseline model

We start with a simple model in which we aim to model the number of flowers/fruits $y_{i,p}$ of each individual $i$ in plot $p$ as a function of its $height$ and neighbourhood crowding $nci$ with a random effect on the plot, such as:

$$y_{i,p} \sim \mathcal Plog (\mu_p + \beta*nci_i)$$
$$\mu_p \sim \mathcal N (\mu, \sigma)$$

### Flowers

```{r baseFitFullFlowers, eval=FALSE}
m_base <- stan_model("model/base.stan")
f_base_full_flowers <- sampling(m_base, mdata_flowers, save_warmup = F, chains = 2, cores = 2, include = F, pars = c("count", "mu_p"))
save(f_base_full_flowers, file = "save/base_full_flowers.Rdata")
```

```{r envOutFullFlowers}
load("save/base_full_flowers.Rdata")
cowplot::plot_grid(
  mcmc_trace(f_base_full_flowers, regex_pars = c("mu", "beta", "sigma")),
  mcmc_intervals(f_base_full_flowers, regex_pars = c("mu", "beta", "sigma")),
  ppc_dens_overlay(y = data$flowers, as.matrix(f_base_full_flowers, pars = "prediction")[1:100, ]),
  nrow = 2
)
```

### Fruits

```{r baseFitFullFruits, eval=FALSE}
m_base <- stan_model("model/base.stan")
f_base_full_fruits <- sampling(m_base, mdata_fruits, save_warmup = F, chains = 2, cores = 2, include = F, pars = c("count", "mu_p"))
save(f_base_full_fruits, file = "save/base_full_fruits.Rdata")
```

```{r envOutFullFruits}
load("save/base_full_fruits.Rdata")
cowplot::plot_grid(
  mcmc_trace(f_base_full_fruits, regex_pars = c("mu", "beta", "sigma")),
  mcmc_intervals(f_base_full_fruits, regex_pars = c("mu", "beta", "sigma")),
  ppc_dens_overlay(y = data$fruits, as.matrix(f_base_full_fruits, pars = "prediction")[1:100, ]),
  nrow = 2
)
```

## Hierarchical model

Let's assume there are two genetically distinct types of individuals that can be differentiated based on their flowering date : early or late flowering. We want to identify individuals with the genetic potential of flowering earlier. Let's p be the probability of a late individual i and 1−p the probability of an early individual.

$$y_{i,p} \sim \mathcal Plog (\mu_p^l + \beta*nci_i)$$
$$\mu_p^l \sim \mathcal N (\mu^l, \sigma)$$
$$l\in{1,2}|z_{i,p}\sim\mathcal B(p)$$

### Flowers

```{r hierFitFullFlowers, eval=FALSE}
m_hier <- stan_model("model/hier.stan")
f_hier_full_flower <- sampling(m_hier, mdata_flowers, save_warmup = F, chains = 2, cores = 2, include = F, pars = c("delta", "count"))
save(f_hier_full_flower, file = "save/hier_full_flowers.Rdata")
```

```{r hierOutFullFlowers}
load("save/hier_full_flowers.Rdata")
cowplot::plot_grid(
  mcmc_trace(f_hier_full_flower, regex_pars = c("mu0", "diff", "beta", "sigma", "p$")),
  mcmc_intervals(f_hier_full_flower, regex_pars = c("mu0", "diff", "beta", "alpha", "gamma", "sigma")),
  ppc_dens_overlay(y = data$flowers, as.matrix(f_hier_full_flower, pars = "prediction")[1:100, ]),
  mcmc_areas(f_hier_full_flower, "p"),
  nrow = 2
)
```

```{r hierCompFullFlowers}
loo_compare(loo(f_base_full_flowers), loo(f_hier_full_flower)) %>%
  kable()
```

### Fruits

```{r hierFitFullFruits, eval=FALSE}
m_hier <- stan_model("model/hier.stan")
f_hier_full_fruits <- sampling(m_hier, mdata_fruits, save_warmup = F, chains = 2, cores = 2, include = F, pars = c("delta", "count"))
save(f_hier_full_fruits, file = "save/hier_full_fruits.Rdata")
```

```{r hierOutFullFruits}
load("save/hier_full_fruits.Rdata")
cowplot::plot_grid(
  mcmc_trace(f_hier_full_fruits, regex_pars = c("mu0", "diff", "beta", "sigma", "p$")),
  mcmc_intervals(f_hier_full_fruits, regex_pars = c("mu0", "diff", "beta", "alpha", "gamma", "sigma")),
  ppc_dens_overlay(y = data$fruits, as.matrix(f_hier_full_fruits, pars = "prediction")[1:100, ]),
  mcmc_areas(f_hier_full_fruits, "p"),
  nrow = 2
)
```

```{r hierCompFullFruits}
loo_compare(loo(f_base_full_fruits), loo(f_hier_full_fruits)) %>%
  kable()
```

## Hierarchical model with covariable

$$y_{i,p} \sim \mathcal Plog (\mu_p^l + \beta*nci_i)$$
$$\mu_p^l \sim \mathcal N (\mu^l, \sigma)$$
$$l\in{1,2}|z_{i,p}\sim\mathcal B(\eta_0 + \gamma*shore_p)$$

### Flowers

```{r hiercovFitFullFlowers, eval=FALSE}
m_hiercov <- stan_model("model/hiercov.stan")
f_hiercov_full_flowers <- sampling(m_hiercov, mdata_flowers, save_warmup = F, chains = 2, cores = 2, include = F, pars = c("delta", "count"))
save(f_hiercov_full_flowers, file = "save/hiercov_full_flowers.Rdata")
```

```{r hiercovOutFullFlowers}
load("save/hiercov_full_flowers.Rdata")
cowplot::plot_grid(
  mcmc_trace(f_hiercov_full_flowers, regex_pars = c("mu0", "diff", "beta", "eta", "gamma", "sigma")),
  mcmc_intervals(f_hiercov_full_flowers, regex_pars = c("eta", "gamma", "mu0", "diff", "beta", "alpha", "gamma", "sigma")),
  ppc_dens_overlay(y = data$flowers, as.matrix(f_hiercov_full_flowers, pars = "prediction")[1:100, ]),
  nrow = 2
)
```

```{r hiercovCompFullFlowers}
loo_compare(loo(f_base_full_flowers), loo(f_hier_full_flower), loo(f_hiercov_full_flowers)) %>% 
  kable()
```

### Fruits

```{r hiercovFitFullFruits, eval=FALSE}
m_hiercov <- stan_model("model/hiercov.stan")
f_hiercov_full_fruits <- sampling(m_hiercov, mdata_fruits, save_warmup = F, chains = 2, cores = 2, include = F, pars = c("delta", "count"))
save(f_hiercov_full_fruits, file = "save/hiercov_full_fruits.Rdata")
```

```{r hiercovOutFullFruits}
load("save/hiercov_full_fruits.Rdata")
cowplot::plot_grid(
  mcmc_trace(f_hiercov_full_fruits, regex_pars = c("mu0", "diff", "beta", "eta", "gamma", "sigma")),
  mcmc_intervals(f_hiercov_full_fruits, regex_pars = c("eta", "gamma", "mu0", "diff", "beta", "alpha", "gamma", "sigma")),
  ppc_dens_overlay(y = data$fruits, as.matrix(f_hiercov_full_fruits, pars = "prediction")[1:100, ]),
  nrow = 2
)
```

```{r hiercovCompFullFruits}
loo_compare(loo(f_base_full_fruits), loo(f_hier_full_fruits), loo(f_hiercov_full_fruits)) %>% 
  kable()
```
