```{r setupfruits, include=FALSE, cache=F}
rm(list = ls()) ; invisible(gc()) ; set.seed(42)
library(knitr)
library(kableExtra)
if(knitr:::is_html_output()) options(knitr.table.format = "html") 
if(knitr:::is_latex_output()) options(knitr.table.format = "latex") 
library(tidyverse)
library(ggfortify)
library(rstan)
library(loo)
library(bayesplot)
theme_set(bayesplot::theme_default())
options(mc.cores = 4)
rstan_options(auto_write = TRUE)
opts_chunk$set(
  echo = F, message = F, warning = F, fig.height = 6, fig.width = 8,
  cache = T, cache.lazy = F)
```

# Flowers

```{r mdatafruits}
data <- read_tsv("data/mangrove_data.txt") %>% 
  dplyr::rename(plot = `Placettes`,
                id = `ID`,
                height = `Hauteur(m)`,
                d1 = `Diametre1(mm)`,
                d2 = `Diametre2(mm)`,
                diameter = `Moyennediam(mm)`,
                branches = `Branches(N)`,
                fruits = `Fruits(N)`,
                flowers = `Fleurs(N)`,
                pflower= `Presencefleurs(1/0)`,
                nci1 = `Competition_1(cm)`,
                nci2 = `Competition_2(cm)`,
                nci3 = `Competition_3(cm)`,
                nci = `Moyennecompetition(cm)`,
                density = `Densite(trees/m2)`,
                date = `Date`,
                time = `Heure`,
                soil = `Typesol`,
                openness = `Ouverture`,
                age = `Agemax(months)`,
                shore = `Distance(fromShore)`) %>% 
  mutate(time = as.numeric(gsub("h", "", time))) %>% 
  dplyr::select(
    id,
    flowers, fruits,
    height, nci, density, plot
  ) %>% 
  mutate(plotNum = as.numeric(as.factor(plot))) %>% 
  mutate_at(c("nci", "density"),
            scale) %>% 
  mutate_at(c("nci", "density"),
            as.vector)
mdata <- list(
  I = nrow(data),
  P = max(data$plotNum),
  y = data$fruits,
  height = data$height,
  nci = data$nci,
  density = data$density,
  plot = data$plotNum
)
```

## Baseline model

We start with a simple model in which we aim to model the number of flowers $fruits_{i,p}$ of each individual $i$ in plot $p$ as a function of its $height$ and neighbourhood crowding $nci$ with a random effect on the plot, such as:

$$fruits_{i,p} \sim \mathcal Plog (\mu_p + \beta*nci_i)$$
$$\mu_p \sim \mathcal N (\mu, \sigma)$$


```{r baseFit2, eval=FALSE}
m_base <- stan_model("model/base.stan")
f_base <- sampling(m_base, mdata, save_warmup = F, chains = 2, cores = 2, include = F, pars = c("count", "mu_p"))
save(f_base, file = "save/base2.Rdata")
```

```{r envOut2}
load("save/base2.Rdata")
cowplot::plot_grid(
  mcmc_trace(f_base, regex_pars = c("mu", "beta", "sigma")),
  mcmc_intervals(f_base, regex_pars = c("mu", "beta", "sigma")),
  ppc_dens_overlay(y = data$fruits, as.matrix(f_base, pars = "prediction")[1:100, ]),
  nrow = 2
)
```

## Hierarchical model

We are now adding an hidden state to the best model. Let's assume there are two genetically distinct types of individuals that can be differentiated based on their flowering date : early or late flowering. We want to identify individuals with the genetic potential of flowering earlier. Let's p be the probability of a late individual i and 1−p the probability of an early individual.

$$fruits_{i,p} \sim \mathcal Plog (\mu_p^l + \beta*nci_i)$$
$$\mu_p^l \sim \mathcal N (\mu^l, \sigma)$$
$$l\in{1,2}|z_{i,p}\sim\mathcal B(p)$$

```{r hierFit2, eval=FALSE}
m_hier <- stan_model("model/hier.stan")
f_hier <- sampling(m_hier, mdata, save_warmup = F, chains = 2, cores = 2, include = F, pars = c("delta", "count"))
save(f_hier, file = "save/hier2.Rdata")
```

```{r hierOut2}
load("save/hier2.Rdata")
cowplot::plot_grid(
  mcmc_trace(f_hier, regex_pars = c("mu0", "diff", "beta", "sigma", "p$")),
  mcmc_intervals(f_hier, regex_pars = c("mu0", "diff", "beta", "alpha", "gamma", "sigma")),
  ppc_dens_overlay(y = data$fruits, as.matrix(f_hier, pars = "prediction")[1:100, ]),
  mcmc_areas(f_hier, "p"),
  nrow = 2
)
```

```{r hierComp2}
loo_compare(loo(f_base), loo(f_hier)) %>%
  kable()
```

## Hierarchical model with covariable

$$fruits_{i,p} \sim \mathcal Plog (\mu_p^l + \beta*nci_i)$$
$$\mu_p^l \sim \mathcal N (\mu^l, \sigma)$$
$$l\in{1,2}|z_{i,p}\sim\mathcal B(\eta_0 + \gamma*shore_p)$$

```{r hiercovFit2, eval=FALSE}
m_hiercov <- stan_model("model/hiercov.stan")
f_hiercov <- sampling(m_hiercov, mdata, save_warmup = F, chains = 2, cores = 2, include = F, pars = c("delta", "count"))
save(f_hiercov, file = "save/hiercov2.Rdata")
```

```{r hiercovOut2}
load("save/hiercov2.Rdata")
cowplot::plot_grid(
  mcmc_trace(f_hiercov, regex_pars = c("mu0", "diff", "beta", "eta", "gamma", "sigma")),
  mcmc_intervals(f_hiercov, regex_pars = c("eta", "gamma", "mu0", "diff", "beta", "alpha", "gamma", "sigma")),
  ppc_dens_overlay(y = data$fruits, as.matrix(f_hiercov, pars = "prediction")[1:100, ]),
  nrow = 2
)
```

```{r hiercovComp2}
loo_compare(loo(f_base), loo(f_hier), loo(f_hiercov)) %>% 
  kable()
```
