```{r setuphiermodel2, include=FALSE, cache=F}
rm(list = ls()) ; invisible(gc()) ; set.seed(42)
library(knitr)
library(kableExtra)
if(knitr:::is_html_output()) options(knitr.table.format = "html") 
if(knitr:::is_latex_output()) options(knitr.table.format = "latex") 
library(tidyverse)
library(ggfortify)
library(rstan)
library(loo)
library(bayesplot)
theme_set(bayesplot::theme_default())
options(mc.cores = 4)
rstan_options(auto_write = TRUE)
opts_chunk$set(
  echo = F, message = F, warning = F, fig.height = 6, fig.width = 8,
  cache = T, cache.lazy = F)
```

# Hierarchical model - Normal

```{r mdatahier2}
data <- read_tsv("data/mangrove_data.txt") %>% 
  dplyr::rename(plot = `Placettes`,
                id = `ID`,
                height = `Hauteur(m)`,
                d1 = `Diametre1(mm)`,
                d2 = `Diametre2(mm)`,
                diameter = `Moyennediam(mm)`,
                branches = `Branches(N)`,
                fruits = `Fruits(N)`,
                flowers = `Fleurs(N)`,
                pflower= `Presencefleurs(1/0)`,
                nci1 = `Competition_1(cm)`,
                nci2 = `Competition_2(cm)`,
                nci3 = `Competition_3(cm)`,
                nci = `Moyennecompetition(cm)`,
                density = `Densite(trees/m2)`,
                date = `Date`,
                time = `Heure`,
                soil = `Typesol`,
                openness = `Ouverture`,
                age = `Agemax(months)`,
                shore = `Distance(fromShore)`) %>% 
  mutate(time = as.numeric(gsub("h", "", time))) %>% 
  dplyr::select(
    id,
    height, diameter,
    pflower,
    nci, density, shore,
    soil, openness,
    plot
  ) %>% 
  mutate(plotNum = as.numeric(as.factor(plot))) %>% 
  mutate_at(c("nci", "density", "shore"),
            scale) %>% 
  mutate_at(c("nci", "density", "shore"),
            as.vector)
mdata_height <- list(
  I = nrow(data),
  S = max(data$soil),
  O = max(data$openness),
  P = max(data$plotNum),
  y = data$height,
  nci = data$nci,
  density = data$density,
  shore = data$shore,
  soil = data$soil,
  openness = data$openness,
  plot = data$plotNum
)
mdata_diameter <- list(
  I = nrow(data),
  S = max(data$soil),
  O = max(data$openness),
  P = max(data$plotNum),
  y = data$diameter,
  nci = data$nci,
  density = data$density,
  shore = data$shore,
  soil = data$soil,
  openness = data$openness,
  plot = data$plotNum
)
```

## Hierarchical model

We are now adding an hidden state to the best model. Let's assume there are two genetically distinct types of individuals that can be differentiated based on their flowering date : early or late flowering. We want to identify individuals with the genetic potential of flowering earlier. Let's p be the probability of a late individual i and 1−p the probability of an early individual.

$$log(y_{i,p}) \sim \mathcal N (\alpha_p^l + \beta_1*nci_i + \beta_2*density_p + \beta_3*shore_p, \sigma)$$
$$\mu_p \sim \mathcal N (\alpha^l, \sigma_{plot})$$

$$l\in{1,2}|z_{i,p}\sim\mathcal B(p)$$

### Height

```{r hierFitHeight, eval=FALSE}
m_hierlog <- stan_model("model2/hierlog.stan")
f_hierlog_height <- sampling(m_hierlog, mdata_height, save_warmup = F, chains = 2, cores = 2, include = F, pars = c("mu_p", "odd"))
save(f_hierlog_height, file = "save/hierlogheight.Rdata")
```

```{r hierOutHeight}
load("save/hierlogheight.Rdata")
cowplot::plot_grid(
  mcmc_trace(f_hierlog_height, regex_pars = c("alpha0", "diff", "beta", "sigma")),
  mcmc_intervals(f_hierlog_height, regex_pars = c("alpha0", "diff", "beta", "alpha", "gamma", "sigma")),
  ppc_dens_overlay(y = data$height, as.matrix(f_hierlog_height, pars = "prediction")[1:100, ]),
  nrow = 2
)
```

### Diameter

```{r hierFitDiameter, eval=FALSE}
m_hierlog <- stan_model("model2/hierlog.stan")
f_hierlog_diameter<- sampling(m_hierlog, mdata_diameter, save_warmup = F, chains = 2, cores = 2, include = F, pars = c("mu_p", "odd"))
save(f_hierlog_diameter, file = "save/hierlogdiameter.Rdata")
```

```{r hierOutDiameter}
load("save/hierlogdiameter.Rdata")
cowplot::plot_grid(
  mcmc_trace(f_hierlog_diameter, regex_pars = c("alpha0", "diff", "beta", "sigma")),
  mcmc_intervals(f_hierlog_diameter, regex_pars = c("alpha0", "diff", "beta", "alpha", "gamma", "sigma")),
  ppc_dens_overlay(y = data$diameter, as.matrix(f_hierlog_diameter, pars = "prediction")[1:100, ]),
  nrow = 2
)
```


